Transmembrane helices are also an overlooked source of major histocompatibility complex class II epitopes and evolutionary more conserved than expected by chance

### 2021-03-26 bbbq_article.tex
Process part of Geert's feedback

While humans only express a limited number of MHC haplotypes, and therefore any individual's immune system detects only a fraction of all possible
peptide fragments, the number of pathogenic peptides that can be detected is increased at the population level, because of the multiple and highly polymorphic MHC genes.
It is believed that this improves immunity at the population level, 
as mutations in a protein that disrupt a particular MHC presentation, 
so-called escape mutations, 
will not affect MHC presentation for all haplotypes (\cite{sommer2005importance}).
[GEERT: I removed the nr of MHC molecules at is is more complex, as the alpha chains coded by the maternal chromosome of HLA-DQ,DR,DP can heterodimeriza with the beta chains coded by the paternal chromosome and vice versa]

@@

Many studies are aimed at determining which peptides are presented in MHC 
and will result in an immune response, 
as this will for instance aid the design of vaccines. 
These studies have led to the development 
of  prediction algorithms 
that allow for very reliable in silico predictions 
of the binding affinities of peptides (\cite{larsen2010identification,schellens2008unanticipated,tang2011genome}).[GEERT: can you put 'very reliable' in more quantitative terms (e.g. with overall more than 90\% accuracy)]

@@

Transmembrane helices (TMHs) are hydrophobic 
as they need to span the hydrophobic lipid bilayer of cellular membranes.
They consist of an alpha helix of on average 23 amino acids in length. TMHs can also be predicted from a protein sequence 
with high accuracy [GEERT: more quantitative?] by bioinformatics approaches (\cite{krogh2001predicting,bianchi2017,kall2004combined,arai2004conpred,jones2007improving,klammer2009metatm,wang2019efficient}). 
TMHs are common structures in the proteins of humans and microbes. 
Different TMH prediction tools estimate
that 15-39\% of the human proteome 
contain at least one TMH (\cite{ahram2006estimation}).

However, the physiological reason why peptides derived from TMHs 
would be presented more often than peptides 
stemming from soluble protein regions is unknown. 
We hypothesized is that the presentation of 
TMH residues is evolutionary selected for, 
because TMHs are less prone to undergo escape mutation. 
One reason to expect such a reduced 
variability (and hence evolutionary conservation) in TMHs, 
is that these are restricted in their evolution 
by the functional requirement to span a lipid bilayer. 
Due to this requirement, 
many of the amino acids genuinely present in TMH 
are limited to the ones with hydrophobic side chains (\cite{jones1994model}).
[FB: Analyzing the occurrence of single nucleotide polymorphisms 
that were predicted to be inside ...][RJCB: Uh, you want to add another
paragraph here?][GEERT: I think it is OK like Richel wrote]
Therefore, we speculated that the TMHs of pathogens 
might have a lower chance to develop escape mutations, 
as many mutations will result in a dysfunctional TMH 
and render the protein inactive.

@@

As a data source, multiple
NCBI (\url{https://www.ncbi.nlm.nih.gov/}) databases were used: the 'dbSNP' (\cite{sherry2001dbsnp}) database,
which contains 650 million 
catalogued non-redundant humane variations (called RefSNPs,
\url{https://www.ncbi.nlm.nih.gov/snp/docs/RefSNP_about/}), and the databases 'gene' (for gene names)
and 'protein' (for proteins sequences). [GEERT: any references for databases gene and protein?]

@@

% \paragraph{Selection of SNPs}
%
SNPs were picked based on ID number, which is linked to their discovery date. To verify that ID number is unrelated to SNP positions the relative positons of the analyzed SNPS in a protein was determined and showed no positional bias, shown in figure \ref{fig:snp_rel_pos}.[GEERT: rood practice is to present the data in a chronological fashion: thus move this to figure 1 (or put in supplement)]

>>>>

While humans only express a limited number of MHC haplotypes, and therefore any individual's immune system detects only a fraction of all possible
peptide fragments, the number of pathogenic peptides that can be detected is increased at the population level, because of the multiple and highly polymorphic MHC genes.
It is believed that this improves immunity at the population level, 
as mutations in a protein that disrupt a particular MHC presentation, 
so-called escape mutations, 
will not affect MHC presentation for all haplotypes (\cite{sommer2005importance}).
% Geert: I removed the number of MHC molecules as is is more complex, 
% as the alpha chains coded by the maternal chromosome 
% of HLA-DQ, DR, DP can heterodimerize with the beta chains coded 
% by the paternal chromosome and vice versa

@@

Many studies are aimed at determining which peptides are presented in MHC 
and will result in an immune response, 
as this will for instance aid the design of vaccines. 
These studies have led to the development 
of  prediction algorithms 
that allow for very reliable in silico predictions 
of the binding affinities of peptides (\cite{larsen2010identification,schellens2008unanticipated,tang2011genome}).
For example, \cite{tang2011genome} found that, 
of the 432 peptides that were predicted to bind to MHC,
86\% were confirmed to do so in vivo.

@@

Transmembrane helices (TMHs) are hydrophobic 
as they need to span the hydrophobic lipid bilayer of cellular membranes.
They consist of an alpha helix of on average 23 amino acids in length. TMHs can also be predicted from a protein sequence 
with high accuracy by bioinformatics approaches (\cite{krogh2001predicting,bianchi2017,kall2004combined,arai2004conpred,jones2007improving,klammer2009metatm,wang2019efficient}),
for example, \cite{jones2007improving} found that,
from 184 TMPs with known topology, 
80\% of the TMH predictions replicated this finding.

TMHs are common structures in the proteins of humans and microbes. 
Different TMH prediction tools estimate
that 15-39\% of the human proteome 
contain at least one TMH (\cite{ahram2006estimation}).
However, the physiological reason why peptides derived from TMHs 
would be presented more often than peptides 
stemming from soluble protein regions is unknown. 
We hypothesized is that the presentation of 
TMH residues is evolutionary selected for, 
because TMHs are less prone to undergo escape mutation. 
One reason to expect such a reduced 
variability (and hence evolutionary conservation) in TMHs, 
is that these are restricted in their evolution 
by the functional requirement to span a lipid bilayer. 
Due to this requirement, 
many of the amino acids genuinely present in TMH 
are limited to the ones with hydrophobic side chains (\cite{jones1994model}).
Therefore, we speculated that the TMHs of pathogens 
might have a lower chance to develop escape mutations, 
as many mutations will result in a dysfunctional TMH 
and render the protein inactive.

@@

As a data source, multiple
NCBI (\url{https://www.ncbi.nlm.nih.gov/}) databases were used: the 'dbSNP' (\cite{sherry2001dbsnp}) database,
which contains 650 million 
catalogued non-redundant humane variations (called RefSNPs,
\url{https://www.ncbi.nlm.nih.gov/snp/docs/RefSNP_about/}), and the databases 'gene' (for gene names, \cite{brown2015gene})
and 'protein' (for proteins sequences, \cite{sayers2010database}).

@@

% \paragraph{Selection of SNPs}
%
SNPs were picked based on ID number, which is linked to their discovery date. To verify that ID number is unrelated to SNP positions the relative positons of the analyzed SNPS in a protein was determined and showed no positional bias, shown in figure \ref{fig:snp_rel_pos}.[GEERT: good practice is to present the data in a chronological fashion: thus move this to figure 1 (or put in supplement)]
